---
title: "GEAN manual"
author: "Baoxing Song"
date: "June 12, 2019"
output:
  pdf_document:
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_depth: 4
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = 'H')
```

# GEAN
Here we provide a solution for INDEL inconsistent alignment problem which could lead to false positive splice sites disturb or ORF-shift predication. And whole genome MSA is all developped basing on the genetic features.
\begin{center}
\includegraphics{doc/alignment.png}
\end{center}
By soving this problem, GEAN could also use to transform the well annotated genetics feature of model species to the genome of other natural variation individuls or phylogenetically nearby species with whole genome avaliable.

The inconsistent alignment problem could affect the function impact annotation of INDELs (non-coding INDEL V.S. ORF-shift INDEL), SNP (non-coding region SNP, coding region SNP), by re-alignment, those variants could be moved to non-coding regions.

## Install
### Dependencies
CPU support avx2
GNU GCC >=6.0 
Cmake >= 3.0
Due to the high computational density of weighted sequence alignment algorithm, GEAN only fullly works on hadware platform with CPU support AVX2 constructions.
Some functions could work on most of hardware plarform. As long as you are not using a very old machine, AVX2 should be valiable.

### Installation
```{bash eval = FALSE}
git clone https://github.com/baoxingsong/GEAN.git
cd GEAN
cmake CMakeLists.txt
make
```
## Usage

### Examples
example on different purpose and using genome with different genome complexity could be found on our [github: https://github.com/baoxingsong/GEAN/tree/master/example](https://github.com/baoxingsong/GEAN/tree/master/example)

```{r engine = 'bash', eval = TRUE}
/home/bs674/software/bin/gean
```
### variant calling
Those functions are designed for whole-genome resequencing variant calling data. It works very for sdi file, which is a very simple file format.<br />
For VCF format, you should make sure there is no heterozygous variant calling result. If your are working heterozygous line, you could do phasing and seperate your variant calling result into two or more VCF files. please make sure you have only those variants records pass quanlity control in the input file.<br />
Since the VCF file format really diverse from different variant calling software, we could not make our software work with all the VCF files, it is highly recommanded to reform yoyr vcf file into sdi file.

#### get pseudo genome sequence using reference genome sequence and variant calling result

```{bash}
/home/bs674/software/bin/gean pseudogeno
```

** -prefix is the prefix of chromosome name for vcf/sdi variant records. Like the chromosome in TAIR10 reference genome is Chr1, Chr2, Chr3, Chr4 and Chr5. While the chromosomes in vcf files from the 1001 genomes project were indicated with 1, 2, 3, 4 and 5.
So `-prefix Chr` should be set to make the software work properly. If this parameter is not set correctly, the software would act as no variant records in the input vcf/sdi file.   

#### liftover reference coordinate to pseudo-genome-sequence
Project/liftover a certain reference genome-sequence coordinate to re-sequencing accession/line pseudo-genome-sequence.
```{bash}
/home/bs674/software/bin/gean lift
```


#### liftover pseudo-genome-sequence coordinate to reference genome sequence
Project/liftover a certain coordinate of re-sequencing accession/line pseudo-genome-sequence to reference genome-sequence.
```{bash}
/home/bs674/software/bin/gean revlift
```


#### liftover reference gff/gtf/gff3 annotation to pseudo-genome-sequence
Inference the gene structure (gtf/gff file) annotation of re-sequencing accession/line by purely coordinate liftover.
```{bash}
/home/bs674/software/bin/gean liftgff
```

#### liftover pseudo-genome-sequence gff/gtf/gff3 annotation to reference genome sequence
Project/liftover the gene structure (gtf/gff file) annotation of re-sequencing accession/line to reference genome-sequence by purely coordinate liftover.
```{bash}
/home/bs674/software/bin/gean revliftgff
```


#### re calling variant calling by align the genic region sequencing and keep the completeness of ORF
Realign the sequence using ZDP algorithm to solve the inconsistent INDEL alignment problem and recall all variants which could cause false positive ORF-state shit predication.
```{bash}
/home/bs674/software/bin/gean reanva
```
 * By ORF-states, this software has following criteria:
     1) Splicing sites is one of motif in "SpliceSites", which is included in the release
     2) The minimum length of intron is larger than a certain value
     3) CDS sequence length is larger than a certain value
     4) The length of CDS sequence is divisible by 3
     5) No premature stop codon
     6) End with end codon
     7) Start with start codon
     The IUPAC Codes of DNA sequence could be well dealt with. 
     The result of ORF-states are included in the CDS sequence

#### extract sequece using genome sequence and annotation file
Extract CDS sequence, C-DNA sequence and protein sequence for each protein-coding transcript. And predict the protein coding potential (termed as ORF-state)
```{bash}
/home/bs674/software/bin/gean gff2seq
```

#### annotate the pseudo-genome-sequence
Transform the reference gene structure annotation to re-sequencing accession/lines with several complementary methods.
```{bash}
/home/bs674/software/bin/gean annowgr
```

#### simulate random variants
Assign a random position for each variant in a variant calling result file, which could be used to compare the different between observed variant calling and random variants.
```{bash}
/home/bs674/software/bin/gean generateRandomSdi
```

### Whole genome wide multiple sequence alignment pipeline
#### cut the (pseudo-)genome sequence of a population of individuals into fragments to perform multiple sequence alignment for each fragment.

```{bash}
/home/bs674/software/bin/gean premsa
```

#### variant calling using multiple sequence alignment
perform variant calling from the multiple sequence alignment of sequence fragments of a population of genome sequences
```{bash}
/home/bs674/software/bin/gean msatosdi
```

### project reference annotation to de novo assemly genome sequence
pipeline to project the reference gene structure annotation to a de novo assembly genome sequence highly similar with the reference genome sequence
#### transgff
liftover reference gene structure annotation to a de novo assembly genome sequence using whole genome sequence alignment.
The result file contains duplication gene annotations records, which might do not compile with other software and could be purified with the following function.
```{bash}
/home/bs674/software/bin/gean transgff
```
 
#### purifygff   
remove those duplication gene structure annotations generated from the transff function
```{bash}
/home/bs674/software/bin/gean purifygff
```

### Acknowledgements
The GEAB team would like to thank all our enthusiastic users who have contacted us with
suggestions to improve the codebase, request new functions, point out bugs, and beta-test the initial
versions of GEAN. Many thanks also to everyone who has used GEAN and
cited the publication – we are glad it has proven useful in your research, and a good citation
record will help us to obtain future funding to keep developing GEAN.

### Citation
If you use GEAN, please cite: <br />
`
Baoxing Song, Qing Sang, Hai Wang, Huimin Pei, Fen Wang and Xiangchao Gan. (2019) A weighted sequence alignment strategy for gene structure annotation lift over from reference genome to a newly sequenced individual. bioRxiv. doi:10.1101/615476
`
